On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-02 deprecated.tex
text in appendix and figures
fixed case done? @_@ very close...

@@

>>>>

\section{Disjointed notes}

\subsection{Branch lengths with $x$ and $y$ known}

In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w')
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Some computer work shows that the first likelihood will always be chosen over the second when maximizing ancestral states, so we focus on that.
The likelihood can be written
$$
\ell = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + log(1+w')
$$
where
$$
\alpha_1 = 1+x^2+y^2+4xy^2+x^2y^2,
\alpha_2 = 1+x^2+y^2-4xy^2+x^2y^2,
\beta = 1+x^2+y^2+x^2y^2,
\gamma = 4xy.
$$
The derivative of this is quadratic in w', with
$$
\ell' = (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2) + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) + w^2(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2),
$$
so we have a closed form solution for $w'$ via the quadratic formula, where
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
and
$$
\hat{w}_1' = (-b + \sqrt{b^2 - 4ac}) / 2a,
\hat{w}_2' = (-b - \sqrt{b^2 - 4ac}) / 2a.
$$
These are all functions of $x$ and $y$.
Let's plan to get these in as simplified a form as possible.
Then, we need cases where either $\hat{w}_1'\in[0,1]$ or $\hat{w}_2'\in[0,1]$, and, if not, we set them equal to their extrema.
Whichever yields a higher value of the likelihood will then be the maximizer.
The interesting case, to me, is that where $\hat{w}_2' \ge 1$, which is the case where we estimate $w'$ to be one.
We can plot this in two variables, as we did before.
Let's see.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{bl-inconsistency}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

\subsection{Below is wrong}

My initial hunch is that $w'$ is always equal to 1, though the empirical fits don't show this.
I argue that $w'=1$ always like so: consider just the last term of the likelihood $\log(1+w')$.
This is obviously maximized when $w'=1$.
For the remaining two terms, $p_{\emptyset} \ge p_{13}$, with equality only when $x=y=0$.
So, excepting the case of infinite branch lengths, the term $\log(1+x^2+y^2+4xyw'+x^2y^2)$ will always be weighted more than $\log(1+x^2+y^2-4xyw'+x^2y^2)$.
Additionally, these two terms can be written as
$$
c_1 \cdot f(x, y, w') + c_2 \cdot f(x, y, -w')
$$
where $c_1 > c_2$.
Clearly $\hat{w}_1'=1$ where
$$
\hat{w}_1' = \arg\max \ f(x, y, w')
$$
and
$\hat{w}_2'=0$ where
$$
\hat{w}_2' = \arg\max \ f(x, y, -w').
$$
I would think that $\hat{w}'=1$ where
$$
\hat{w}_2' = \arg\max \ c_1 \cdot f(x, y, w') + c_2 \cdot f(x, y, -w').
$$
This is clearly true if $f$ is linear in $w'$.
Here, though, $f$ is $log(a+b*w')$.
We have ranges $a\in[1,4]$ and $b\in[0,4]$ where $a$ increases quartically with $x$ and $y$ while $b$ increases quadratically.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t^*)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t^*)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$     &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

### 2017-10-02 joint_inf.tex
fixed case done? @_@ very close...

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{analytic-inconsistency}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

>>>>

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{analytic-inconsistency-legend}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the top right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w')
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}

\subsection{Analytic}

In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w')
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
At the risk of sounding circular, if $\hat{w}'$ is estimated to be one, then the first likelihood will be chosen, so the region in which that likelihood estimates $\hat{w}'$ to be one will always do so; let's focus on the first likelihood then.
The likelihood can be written
$$
\ell = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma = 4xy.
$$
The derivative of this is quadratic in w', with
$$
\ell' = (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2) + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) + w^2(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2),
$$
so we have a closed form solution for $w'$ via the quadratic formula, where
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
and
$$
\hat{w}_1' = (-b + \sqrt{b^2 - 4ac}) / 2a,
$$
$$
\hat{w}_2' = (-b - \sqrt{b^2 - 4ac}) / 2a.
$$
These are all functions of $x$ and $y$.
Let's plan to get these in as simplified a form as possible.
Then, we need cases where either $\hat{w}_1'\in[0,1]$ or $\hat{w}_2'\in[0,1]$, and, if not, we set them equal to their extrema.
Whichever yields a higher value of the likelihood will then be the maximizer.
The interesting case, to me, is that where $\hat{w}_2' \ge 1$, which is the case where we guarantee $w'$ will be one.
We can plot this in two variables, as we did before.
Let's see.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{bl-inconsistency}
\caption{Region of inconsistency with $x$ and $y$ fixed}
\label{fig:bl-inconsistency}
\end{figure}

\subsection{Numerical}

When computing the global maximum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
We find that the estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.
